Smooth and Persistent Forecasts of German GDP\\ Balancing Accuracy and Stability\thanks{We would like Barbara Rossi and the participants at the Vienna Forecasting Workshop 2025 for their comments and suggestions.}

### 2025-08-09 Archive IWH Simon paper/10.08.2025/paper.tex
up

We can now extend the SSA criterion \ref{crit1} to a multivariate setting. Let $\mathbf{X}_t$ (of dimension $t\times n$) denote a set of $n$ explanatory series $\mathbf{x}_{1t},...,\mathbf{x}_{nt}$, with observations $\mathbf{x}_{it}=(x_{it},...,x_{i1})'$. Let $\mathbf{z}_t=(z_{t},...,z_1)'$  denote a target series, which typically lies outside the linear space spanned by $\mathbf{X}_t$  (and is unknown at time $t$). In this framework, $\mathbf{X}_t$ corresponds to a matrix of selected economic indicators for the German economy, with $n=5$, and $z_{it}$, i=1,...,5,   are the outputs of a two-sided HP(160) filter applied to doubly-infinite extensions $(x_{ik})$, $|k|<\infty$, of the indicators, including GDP. In a later step, devised in Section \ref{sec:mSSA_component}, the (filtered) predictors $y_{it}(h)$ of $z_{it+h}$, with $h\geq 0$, will be used to setup up a predictor for GDP.%Since $z_{it}$  depends on future observations $x_{it+1},x_{it+2},...$, the prediction task involves `tracking' $z_{it}$  using the estimate $y_{it}=\sum_{j=1}^ny_{ijt}$, where $y_{ijt}=\mathbf{b}_{ij}~'\mathbf{x}_{jt}=\sum_{k=0}^{L} b_{ijk}x_{j,t-k}$ are the outputs of a multivariate (causal or one-sided) filter $\mathbf{B}_i^f$, with columns $\mathbf{b}_{ij}$ of fixed length $L$, $j=1,...,5$, assigning weights to the last $L$ observations of the $j$-th indicator $\mathbf{x}_{jt}$ (the fixed-length assumption is merely used for ease of exposition).
%For the sake of clarity, we now omit the index $i$ for all variables, assuming $z_t=z_{i_0t}$ for some fixed $i_0$.

To introduce the M-SSA framework, start by assuming that  $\mathbf{X}_t=\boldsymbol{\epsilon}_t$ is an $n$-dimensional white noise sequence with a full-rank variance-covariance matrix $\boldsymbol{\Sigma}$.

@@

Here, $\boldsymbol{\gamma}_{h}$ and $\mathbf{b}$ are matrices of dimension $(L\cdot n)\times n$: the $i$-th column, $i=1,...,n$, collects the filter weights of the $i$-th target (or the $i$-th predictor) and the corresponding filter weights assigned to the $j=1,...,n$ series are stacked in the (long) column vector. The $i$-th column vector of $\boldsymbol{\gamma}_{h}$ (or $\mathbf{b}$) can then be applied to the $L\cdot n$-dimensional $\boldsymbol{\epsilon}_{\cdot t}$, the data. Specifically, we define $y_{ijt} := \mathbf{b}_{ij}'\boldsymbol{\epsilon}_{jt}$, which allows us to express $y_{it} = \mathbf{b}_i'\boldsymbol{\epsilon}_{\cdot t} = \sum_{j=1}^n y_{ijt}$ and $\mathbf{y}_t = \mathbf{b}'\boldsymbol{\epsilon}_{\cdot t}$. Through orthogonal projection, under the white noise assumption, the MSE predictor is derived as $\hat{\mathbf{y}}_{t,MSE} = \boldsymbol{\gamma}_{h}'\boldsymbol{\epsilon}_{\cdot t}$, with components $\hat{y}_{it,MSE}$, $i=1,...,n$. Consider the Kronecker products defined as $\tilde{\mathbf{I}}:=\boldsymbol{\Sigma}\otimes\mathbf{I}_{L\times L}$ and $\tilde{\mathbf{M}}:=\boldsymbol{\Sigma}\otimes\mathbf{M}$, of the $n\times n$ dimensional matrix $\boldsymbol{\Sigma}$, the $L\times L$ dimensional identity  $\mathbf{I}_{L\times L}$ and the autocovariance generating matrix $\mathbf{M}$. Under the white noise assumption, we can express the following expectations: 

@@

E[z_{it+h}y_{it}]&=&E[\hat{y}_{it,MSE}y_{it}]=\boldsymbol{\gamma}_{i\cdot h}'\tilde{\mathbf{I}}\mathbf{b}_{i}\nonumber\\%\label{moment1}\\
E[\hat{z}_{ith}^2]&=&\boldsymbol{\gamma}_{i\cdot h}'\tilde{\mathbf{I}}\boldsymbol{\gamma}_{i\cdot h}\nonumber\\%\label{moment2}\\

@@

E[y_{it-1}y_{it}]&=&\mathbf{b}_{i}'\tilde{\mathbf{M}}\mathbf{b}_{i}.\nonumber

@@

The variance of $z_{it}$ is given by 
E[z_{it}^2]=\sum_{|k|<\infty}\boldsymbol{\gamma}_{i k}'\boldsymbol{\Sigma}\boldsymbol{\gamma}_{i k},
where $\boldsymbol{\gamma}_{i k}:=(\gamma_{i1k},...,\gamma_{ink})'$. Consequently, we propose the following multivariate M-SSA criterion for $i=1,...,n$: 

@@

for $i=1,...,n$, where we assume an arbitrary normalization, $l=1$, in the length constraint. In analogy to the univariate case, the objective function is then proportional to the target correlation between the predictor and the $i$-th target. Furthermore, the criterion relies on affine and quadratic functions of $\mathbf{b}_i$, facilitating the numerical optimization. Finally, the HT constraint regulates the first-order ACF or, equivalently, the HT of the predictor.\\

@@

 Consequently, the subsequent analysis employs the M-SSA framework to construct a `smooth' predictor $y_t$ for forecasting $z_{t+h}$. In particular, it is shown that controlling the zero-crossing rate within the M-SSA framework has a quantifiable effect on the hit rate and false positive rate.

@@

Figure \ref{bk_gammak} displays the coefficients of the multivariate nowcasts, where M-SSA's increased smoothing reduces the maximum amplitudes of the filter weights and slows their rate of decay at longer lags. The resulting filtered series, presented in Figure \ref{mssa_msse_now}, are standardized to enable direct comparisons. The M-SSA nowcast is slightly smoother than that of the M-MSE, exhibiting approximately one-third fewer sign-changes. Fig.\ref{mssa_msse_zc} compares both nowcasts in relation to the financial crisis, highlighting that the classic (MSE-based) design tends to produce unwanted `noisy' zero-crossings at (or in the vicinity of) the transitions between recessions and expansions. Further, Table \ref{corhtnow} compares sample target correlations and HTs of the two nowcasts. Increased smoothness, i.e. sample holding time increases from 3.7 to 6.5 quarters, achieved by the M-SSA comes at the cost of a slight reduction (0.68 vs 0.69) in target correlation. This highlights the inherent trade-off—often referred to as the prediction dilemma—between smoothness (zero-crossing rate) and predictive accuracy (target correlation) within the M-SSA criterion. 
The results confirm that the smoothing constraint in M-SSA
% incorporated into the multivariate extension of criterion \eqref{critssa} effectively manages the sign-changes of the predictor: sample estimates of the holding times align with the expected values derived from Eq.~\eqref{ht}.\footnote{Effective convergence of sample estimates to their theoretical expected values —assuming the true data-generating process is known— can be validated using the M-SSA package. This validation is typically performed through analyses of very long samples of simulated (multivariate) data, ensuring the consistency of the estimation procedures under idealized conditions.} An 
increases by $50\%$ the HT relative to that of the M-MSE predictor but has only a modest effect on the target correlation, suggesting that this trade-off could be beneficial in our application
% , as confirmed by a subsequent analysis of hit rate and false alarm rate
. %Going forward, we now omit the M-MSE, as it is a special case of the more general M-SSA framework.

>>>>

We can now extend the SSA criterion \ref{crit1} to a multivariate setting. Let $\mathbf{X}_t$ (of dimension $t\times n$) denote a set of $n$ explanatory series $\mathbf{x}_{1t},...,\mathbf{x}_{nt}$, with observations $\mathbf{x}_{it}=(x_{it},...,x_{i1})'$. Let $\mathbf{Z}_t$, of dimension $t\times m$ (in our case $m=n$) a set of $m$ target series $\mathbf{z}_{it}=(z_{it},...,z_{i1})'$, $i=1,...m$, which typically lie outside the linear space spanned by $\mathbf{X}_t$  (and are unknown at time $t$). In this framework, $\mathbf{X}_t$ represents a matrix of selected economic indicators for the German economy, with $n=5$. The targets $z_{it}$, i=1,...,5,   are obtained by applying a two-sided HP(160) filter to the doubly-infinite extensions $(x_{ik})$, $|k|<\infty$, of these indicators, including GDP. In a subsequent step, as delineated in Section \ref{sec:mSSA_component}, the (filtered) predictors $y_{hit}$ of $z_{it+h}$, with $h\geq 0$,  will be utilized to construct predictors for GDP. Unlike the approach in the previous section, these predictors will be based on multivariate M-SSA filters rather than univariate HP-C designs. %Since $z_{it}$  depends on future observations $x_{it+1},x_{it+2},...$, the prediction task involves `tracking' $z_{it}$  using the estimate $y_{it}=\sum_{j=1}^ny_{ijt}$, where $y_{ijt}=\mathbf{b}_{ij}~'\mathbf{x}_{jt}=\sum_{k=0}^{L} b_{ijk}x_{j,t-k}$ are the outputs of a multivariate (causal or one-sided) filter $\mathbf{B}_i^f$, with columns $\mathbf{b}_{ij}$ of fixed length $L$, $j=1,...,5$, assigning weights to the last $L$ observations of the $j$-th indicator $\mathbf{x}_{jt}$ (the fixed-length assumption is merely used for ease of exposition).
%For the sake of clarity, we now omit the index $i$ for all variables, assuming $z_t=z_{i_0t}$ for some fixed $i_0$.

To introduce the M-SSA framework, we begin by assuming that  $\mathbf{X}_t=\boldsymbol{\epsilon}_t$ is an $n$-dimensional white noise sequence with a full-rank variance-covariance matrix $\boldsymbol{\Sigma}$.

@@

Here, $\boldsymbol{\gamma}_{h}$ (or $\mathbf{b}$) is a matrix of dimension $(L\cdot n)\times n$. The $i$-th column, $i=1,...,n$, contains the filter weights associated with the $i$-th target (or $i$-th predictor), while the filter weights corresponding to the $j=1,...,n$ series (indicators) are stacked into column vectors of length $L\cdot  n$. The $i$-th column vector of $\boldsymbol{\gamma}_{h}$ (or $\mathbf{b}$) can be applied to the $L\cdot n$-dimensional $\boldsymbol{\epsilon}_{\cdot t}$, which stacks all $n$ indicators into a single long data vector. Specifically, we define 
y_{ijt} := \mathbf{b}_{ij}'\boldsymbol{\epsilon}_{jt},
which allows us to express the $i$-th predictor as:
y_{it} = \mathbf{b}_i'\boldsymbol{\epsilon}_{\cdot t} = \sum_{j=1}^n y_{ijt}.
Similarly, the multivariate predictor vector can be written as:
\mathbf{y}_t = \mathbf{b}'\boldsymbol{\epsilon}_{\cdot t}
and the same applies to the target variables. Note that, for setting up the optimization criterion, it is not necessary to utilize the doubly infinite expressions of the target(s), as illustrated below. Under the white noise assumption, the orthogonal projection leads to the mean-squared error (MSE) predictor expressed as:
\hat{\mathbf{y}}_{t,MSE} = \boldsymbol{\gamma}_{h}'\boldsymbol{\epsilon}_{\cdot t},
whose components are $\hat{y}_{it,MSE}$, $i=1,...,n$. We define the Kronecker products: 
\tilde{\mathbf{I}}:=\boldsymbol{\Sigma}\otimes\mathbf{I}_{L\times L},\textrm{ ~and~} \tilde{\mathbf{M}}:=\boldsymbol{\Sigma}\otimes\mathbf{M},
where $\boldsymbol{\Sigma}$ is the variance-covariance matrix of the (white noise) data,  $\mathbf{I}_{L\times L}$ is the  $L\times L$ identity matrix  and  $\mathbf{M}$ is the autocovariance generating matrix introduced earlier. Under the white noise assumption, we can express the following expectations: 

@@

E[z_{it+h}y_{it}]&=&\boldsymbol{\gamma}_{i\cdot h}'\tilde{\mathbf{I}}\mathbf{b}_{i}\nonumber\\%\label{moment1}\\
%\label{moment2}\\

@@

E[y_{it-1}y_{it}]&=&\mathbf{b}_{i}'\tilde{\mathbf{M}}\mathbf{b}_{i},\nonumber

@@

where, for notational simplicity, we assume $\sigma^2=1$. 
%The variance of $z_{it}$ is given by 
%\[
%E[z_{it}^2]=\sum_{|k|<\infty}\boldsymbol{\gamma}_{i %k}'\boldsymbol{\Sigma}\boldsymbol{\gamma}_{i k},
%\]
%where $\boldsymbol{\gamma}_{i k}:=(\gamma_{i1k},...,\gamma_{ink})'$ is the weight assigned to the $n$ indicators at lag $k$. 
Consequently, we propose the following multivariate M-SSA criterion for $i=1,...,n$: 

@@

for $i=1,...,n$, where we assume an arbitrary normalization, $l=1$, in the length constraint $\mathbf{b}_{i}'\tilde{\mathbf{I}}\mathbf{b}_{i}=1$. If explicitly necessary, the arbitrary scaling $l=1$ can be revised and optimized separately at a later stage, after an optimal `unity-length' M-SSA predictor has been determined. In analogy to the univariate case, the objective function is proportional to the target correlation $\rho(z_i,y_i)$ between the $i$-th predictor and the $i$-th target.  Moreover, the criterion involves affine and quadratic forms of $\mathbf{b}_i$, which simplifies the process of numerical optimization. Finally, the HT constraint $\mathbf{b}_{i}'\tilde{\mathbf{M}}\mathbf{b}_{i}=\rho_i$, where we used $l=1$, serves to control the first-order ACF or, equivalently, the HT of the predictor.  \\

@@

The subsequent analysis employs the M-SSA framework to construct `smooth' predictors for forecasting $z_{it+h}$ as well as for forecasting the original (unfiltered) GDP. In particular, it is shown that controlling the zero-crossing rate within the M-SSA framework has a quantifiable effect on the hit rate and false positive rate.

@@

Figure \ref{bk_gammak} displays the coefficients of the multivariate nowcasts, where M-SSA's increased smoothing reduces the maximum amplitudes of the filter weights and slows their rate of decay at longer lags. The resulting filtered series, presented in Figure \ref{mssa_msse_now}, are standardized to enable direct comparisons. The M-SSA nowcast is slightly smoother than that of the M-MSE, exhibiting approximately one-third fewer sign-changes. Fig.\ref{mssa_msse_zc} compares both nowcasts in relation to the financial crisis, highlighting that the classic (MSE-based) design tends to produce unwanted `noisy' zero-crossings at (or in the vicinity of) the transitions between recessions and expansions. Further, Table \ref{corhtnow} compares sample target correlations and HTs of the two nowcasts. Increased smoothness, i.e. sample holding time increases from 3.7 to 6.5 quarters, achieved by the M-SSA comes at the cost of a slight reduction (0.68 vs 0.69) in target correlation. This highlights the inherent trade-off—often referred to as the prediction dilemma—between smoothness (zero-crossing rate) and predictive accuracy (target correlation) within the M-SSA criterion. 
The results confirm that the smoothing constraint in M-SSA
% incorporated into the multivariate extension of criterion \eqref{critssa} effectively manages the sign-changes of the predictor: sample estimates of the holding times align with the expected values derived from Eq.~\eqref{ht}.\footnote{Effective convergence of sample estimates to their theoretical expected values —assuming the true data-generating process is known— can be validated using the M-SSA package. This validation is typically performed through analyses of very long samples of simulated (multivariate) data, ensuring the consistency of the estimation procedures under idealized conditions.} An 
increases by $50\%$ the HT relative to that of the M-MSE predictor\footnote{The convergence of the sample HT to the effectively imposed $HT_1$  can be verified in a simulation study, utilizing the M-SSA package.} but has only a modest effect on the target correlation, suggesting that this trade-off could be beneficial in our application
% , as confirmed by a subsequent analysis of hit rate and false alarm rate
. %Going forward, we now omit the M-MSE, as it is a special case of the more general M-SSA framework.
